High-level cognition during story listening is reflected in high-order dynamic correlations in neural activity patterns

### 2020-10-23 paper/main.tex
took a complete pass through paper, updated bibliography, reorganized sections

Studies of the neural code to date have primarily focused on
univariate or multivariate neural patterns~\citep[for review
see][]{NormEtal06}, or (more recently) on patterns of dynamic
first-order correlations~\citep[i.e., interactions between pairs of
brain structures;][]{MannEtal18, FongEtal19, LuriEtal18, PretEtal17,
  ZouEtal19, DemeEtal19}.  We wondered what the future of this line of
work might hold.  For example, is the neural code mediated by
higher-order interactions between brain structures~\citep[e.g.,
see][]{ReimEtal17}?  Second-order correlations reflect
\textit{homologous} patterns of correlation.  In other words, if the
dynamic patterns of correlations between two regions, $A$ and $B$, are
similar to those between two other regions, $C$ and $D$, this would be
reflected in the second-order correlations between ($A$--$B$) and
($C$--$D$).  In this way, second-order correlations identify
similarities and differences between subgraphs of the brain's
connectome.  Analogously, third-order correlations reflect homologies
between second-order correlations-- i.e., homologous patterns of
homologous interactions between brain regions.  More generally,
higher-order correlations reflect homologies between patterns of
lower-order correlations.  We can then ask: which ``orders'' of
interaction are most reflective of high-level cognitive processes?

@@

\section*{Methods}
Our general approach to efficiently estimating high-order dynamic
correlations comprises four general steps (Fig.~\ref{fig:methods}).
First, we derive a kernel-based approach to computing dynamic pairwise
correlations in a $T$ (timepoints) by $K$ (features) multivariate
timeseries, $\mathbf{X}_0$.  This yields a $T$ by $\mathcal{O}(K^2)$
matrix of dynamic correlations, $\mathbf{Y}_1$, where each row
comprises the upper triangle and diagonal of the correlation matrix at a single
timepoint, reshaped into a row vector (this reshaped vector is
$\left( \frac{K^2 - K}{2} + K \right)$-dimensional).  Second, we apply a dimensionality
reduction step to project the matrix of dynamic correlations back onto
a $K$-dimensional space.  This yields a $T$ by $K$ matrix,
$\mathbf{X}_1$, that reflects an approximation of the dynamic
correlations reflected in the original data.  Third, we use repeated
applications of the kernel-based dynamic correlation step to
$\mathbf{X}_n$ and the dimensionality reduction step to the resulting
$\mathbf{Y}_{n+1}$ to estimate high-order dynamic correlations.  Each
application of these steps to a $T$ by $K$ time series $\mathbf{X}_n$
yields a $T$ by $K$ matrix, $\mathbf{X}_{n+1}$, that reflects the
dynamic correlations between the columns of $\mathbf{X}_n$.  In this
way, we refer to $n$ as the \textit{order} of the timeseries, where
$\mathbf{X}_0$ (order 0) denotes the original data and $\mathbf{X}_n$
denotes (approximated) $n^\mathrm{th}$-order dynamic correlations
between the columns of $\mathbf{X}_0$.  Finally, we use a
cross-validation--based decoding approach to evaluate how well
information contained in a given order (or weighted mixture of orders)
may be used to decode relevant cognitive states.  If including a given
$\mathbf{X}_n$ in the feature set yields higher classification
accuracy on held-out data, we interpret this as evidence that the
given cognitive states are reflected in patterns of
$n^\mathrm{th}$-order correlations.  All of the code used to produce
the figures and results in this manuscript, along with links to the
corresponding datasets, may be found at
\href{https://github.com/ContextLab/timecorr-paper}{github.com/ContextLab/timecorr-paper}.
In addition, we have released a Python toolbox for computing dynamic
high-order correlations in timeseries data; our toolbox may be found
at
\href{https://timecorr.readthedocs.io/}{timecorr.readthedocs.io}.

@@

Given a kernel function $\kappa_t(\cdot)$ for timepoint $t$,
evaluated at timepoints $\tau \in \left[ 1, ..., T \right]$, we
can update the static correlation formula in Equation~\ref{eqn:corr}
to estimate the \textit{instantaneous correlation} at timepoint $t$:
\begin{align}
  \mathrm{timecorr}_{\kappa_t}\left(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)\right) &= \frac{\sum_{\tau=1}^T \left( \mathbf{X}(\tau, i) -
                                       \widetilde{\mathbf{X}}_{\kappa_t}(\cdot,
                                                                                        i) \right)
                                 \left( \mathbf{X}(\tau, j) -
                                        \widetilde{\mathbf{X}}_{\kappa_t}(\cdot,
                                                                                        j)\right)}
              {\sqrt{\sum_{\tau=1}^T
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot,
                                                                                        i))
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, j))}},~\mathrm{where}\label{eqn:timecorr}\\
  \widetilde{\mathbf{X}}_{\kappa_t}(\cdot, k) &= \sum_{\tau=1}^T
                       \kappa_t(\tau)\mathbf{X}(\tau, k),\\
  \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, k)) &= \sum_{\tau=1}^T
                                                  \left(
                                                  \mathbf{X}(\tau, k) -
                            \widetilde{\mathbf{X}}_{\kappa_t}(\cdot, k) \right)^2.
\end{align}
Here
$\mathrm{timecorr}_{\kappa_t}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot,
j))$ reflects the correlation at time $t$ between columns $i$ and $j$
of $\mathbf{X}$, estimated using the kernel $\kappa_t$.  We evaluate
Equation~\ref{eqn:timecorr} in turn for each pair of columns in
$\mathbf{X}$ and for kernels centered on each timepoint in the
timeseries, respectively, to obtain a $T$ by $K$ by $K$ timeseries of
dynamic correlations, $\mathbf{Y}$.  For convenience, we then reshape
the upper triangles and diagonals of each timepoint's correlation matrix into a row
vector to obtain an equivalent $T$ by $\left( \frac{K^2 - K}{2} + K \right)$ matrix.

\subsubsection*{Dynamic inter-subject functional connectivity (DISFC)}
Equation~\ref{eqn:timecorr} provides a means of taking a single
observation matrix, $\mathbf{X}_n$ and estimating the dynamic
correlations from moment to moment, $\mathbf{Y}_{n+1}$.  Suppose that
one has access to a set of multiple observation matrices that reflect
the same phenomenon.  For example, one might collect neuroimaging data
from several experimental participants, as each participant performs
the same task (or sequence of tasks).  Let $\mathbf{X}_n^1$,
$\mathbf{X}_n^2$, ..., $\mathbf{X}_n^P$ reflect the $T$ by $K$
observation matrices ($n = 0$) or reduced correlation matrices
($n > 0$) for each of $P$ participants in an experiment.  We can use
\textit{inter-subject functional connectivity}~\citep[ISFC;
][]{SimoEtal16} to compute the stimulus-driven correlations reflected
in the multi-participant dataset at a given timepoint $t$ using:
\begin{align}
\bar{\mathbf{C}}(t) = M\left(R\left(\frac{1}{2P} \sum_{p=1}^P
  Z\left(\mathbf{Y}_{n+1}^p(t)\right)^\top + Z\left(\mathbf{Y}_{n+1}^p(t)\right)\right)\right),\label{eqn:disfc}
\end{align}
where $M$ extracts and vectorizes the upper triangle and diagonal of a symmetric
matrix, $Z$ is the Fisher $z$-transformation~\citep{Zar10}:
\begin{align}
Z(r) = \frac{\log(1+r) - \log(1-r)}{2},
\end{align}
$R$ is the inverse of $Z$:
\begin{align}
R(z) = \frac{\exp(2z - 1)}{\exp(2z + 1)},
\end{align}
and $\mathbf{Y}_{n+1}^p(t)$ denotes the correlation matrix at timepoint $t$
(Eqn.~\ref{eqn:timecorr}) between each column of $\mathbf{X}_n^p$ and each
column of the average $\mathbf{X}_n$ from all \textit{other}
participants, $\bar{\mathbf{X}}_n^{ \backslash p}$:
\begin{align}
  \bar{\mathbf{X}}_n^{ \backslash p} = \frac{1}{P-1}\sum_{q \in
  \backslash p} \mathbf{X}_n^q,
\end{align}
where $\backslash p$ denotes the set of all participants other than
participant $p$. In this way, the $T$ by $\left( \frac{K^2 - K}{2} + K \right)$ DISFC
matrix $\bar{\mathbf{C}}$ provides a time-varying extension of the ISFC
approach developed by \cite{SimoEtal16}.

@@

Applying dimensionality reduction algorithms to $\mathbf{Y}$ yields an
$\mathbf{X}$ whose columns reflect weighted combinations (or nonlinear
transformations) of the original columns of $\mathbf{Y}$.  This has
two main consequences.  First, with each repeated dimensionality
reduction, the resulting $\mathbf{X}_n$ has lower and lower fidelity
(with respect to what the ``true'' $\mathbf{Y}_n$ might have looked
like without using dimensionality reduction to maintain scalability).
In other words, computing $\mathbf{X}_n$ is a lossy operation.
Second, whereas each column of $\mathbf{Y}_n$ may be mapped
directly onto specific pairs of columns of $\mathbf{X}_{n-1}$, the
columns of $\mathbf{X}_n$ reflect weighted combinations and/or
nonlinear transformations of the columns of $\mathbf{Y}_n$.  Many
dimensionality reduction algorithms are invertible (or approximately
invertible).  However, attempting to map a given $\mathbf{X}_n$ back
onto the original feature space of $\mathbf{X}_0$ will usually require
$\mathcal{O}(TK^{2n})$ space and therefore becomes intractable
as $n$ or $K$ grow large.

\subsubsection*{Graph measure approaches to computing
  $\mathbf{X}_n$}
The above dimensionality reduction approaches to approximating a given
$\mathbf{Y}_n$ with a lower-dimensional $\mathbf{X}_n$ preserve a
(potentially recombined and transformed) mapping back to the original
data in $\mathbf{X}_0$.  We also explore graph measures that instead
characterize each feature's relative \textit{position} in the broader
network of interactions and connections.  To illustrate the
distinction between the two general approaches we explore, suppose a
network comprises nodes $A$, $B$, and $C$.  If $A$ and $B$ exhibit
uncorrelated activity patterns, the functional connection
(correlation) between them will be (by definition) close to 0.
However, if $A$ and $B$ each interact with $C$ in similar ways, we
might attempt to capture those similarities using a measure that
reflects how $A$ and $B$ interact with \textit{other} members of the
network.

@@

Because $\mathbf{X}_n$ has the same shape as the original data
$\mathbf{X}_0$, approximating $\mathbf{Y}_n$ with a lower-dimensional
$\mathbf{X}_n$ enables us to estimate high-order dynamic correlations
in a scalable way.  Given a $T$ by $K$ input matrix, the output of
Equation~\ref{eqn:timecorr} requires $\mathcal{O}(TK^2)$ space to
store.  Repeated applications of Equation~\ref{eqn:timecorr} (i.e.,
computing dynamic correlations between the columns of the outputted
dynamic correlation matrix) each require exponentially more space; in
general the $n^\mathrm{th}$-order dynamic correlations of a $T$ by $K$
timeseries occupies $\mathcal{O}(TK^{2n})$ space.  However, when we
approximate or summarize the output of Equation~\ref{eqn:timecorr} with a $T$ by
$K$ matrix (as described above), it becomes feasible to compute even
very high-order correlations in high-dimensional data.  Specifically,
approximating the $n^\mathrm{th}$-order dynamic correlations of a $T$
by $K$ timeseries requires only $\mathcal{O}(TK^2)$ additional space--
the same as would be required to compute first-order dynamic
correlations. In other words, the space required to store $n+1$
multivariate timeseries reflecting up to $n^\mathrm{th}$ order
correlations in the original data scales linearly with $n$ using our
approach (Fig.~\ref{fig:methods}).

\subsection*{Data}
We examined two types of data: synthetic data and human functional
neuroimaging data.  We constructed and leveraged the synthetic data to
evaluate our general approach~\citep[for a related validation
approach see][]{ThomEtal18}.  Specifically, we tested how well
Equation~\ref{eqn:timecorr} could be used to recover known dynamic
correlations using different choices of kernel ($\kappa$;
Fig.~\ref{fig:kernels}), for each of several synthetic datasets that
exhibited different temporal properties. We also used this synthetic
data to simulate higher-order correlations and tested how well Equation~\ref{eqn:timecorr}
could recover these correlations using the best kernel from the
previous synthetic data analysis.  We applied our approach to a
functional neuroimaging dataset to test the hypothesis that ongoing
cognitive processing is reflected in high-order dynamic correlations.
We used an across-participant classification test to estimate whether
dynamic correlations of different orders contain information about
which timepoint in a story participants were listening to.

>>>>

Studies of the neural code to date have primarily focused on
univariate or multivariate neural patterns~\citep[for review
see][]{NormEtal06}, or (more recently) on patterns of dynamic
first-order correlations~\citep[i.e., interactions between pairs of
brain structures;][]{MannEtal18, FongEtal19, LuriEtal18, PretEtal17,
  ZouEtal19, DemeEtal19}.  What might the future of this line of work
hold?  For example, is the neural code implemented through
higher-order interactions between brain structures~\citep[e.g.,
see][]{ReimEtal17}?  Second-order correlations reflect
\textit{homologous} patterns of correlation.  In other words, if the
dynamic patterns of correlations between two regions, $A$ and $B$, are
similar to those between two other regions, $C$ and $D$, this would be
reflected in the second-order correlations between ($A$--$B$) and
($C$--$D$).  In this way, second-order correlations identify
similarities and differences between subgraphs of the brain's
connectome.  Analogously, third-order correlations reflect homologies
between second-order correlations-- i.e., homologous patterns of
homologous interactions between brain regions.  More generally,
higher-order correlations reflect homologies between patterns of
lower-order correlations.  We can then ask: which ``orders'' of
interaction are most reflective of high-level cognitive processes?

@@

\section*{Results}
We sought to understand whether high-level cognition is reflected in
dynamic patterns of high-order correlations.  To that end, we
developed a computational framework for estimating the dynamics of
stimulus-driven high-order correlations in multivariate timeseries
data (see \textit{Dynamic inter-subject functional connectivity
  (DISFC)} and \textit{Dynamic higher-order correlations}).  We
evaluated the efficacy of this framework at recovering known patterns
in several synthetic datasets (see \textit{Synthetic data: simulating
  dynamic first-order correlations} and \textit{Synthetic data:
  simulating dynamic higher-order correlations}).  We then
applied the framework to a public fMRI dataset collected as
participants listened to an auditorally presented story, listened to a
temporally scrambled version of the story, or underwent a resting
state scan (see \textit{Functional neuroimaging data collected during
  story listening}).  We used the relative decoding accuracies of
classifiers trained on different sets of neural features to estimate
which types of features reflected ongoing cognitive processing.

\subsection*{Recovering known dynamic correlations from synthetic
  data}
\subsubsection*{Recovering dynamic first-order correlations}
We generated synthetic datasets that differed in how the underlying first-order
correlations changed over time.  For each dataset, we applied
Equation~\ref{eqn:timecorr} with a variety of kernel shapes and
widths.  We assessed how well the true underlying correlations at each
timepoint matched the recovered correlations
(Fig.~\ref{fig:synthetic}).  For every kernel and dataset we tested,
our approach recovered the correlation dynamics we embedded into the
data.  However, the quality of these recoveries varied across
different synthetic datasets in a kernel-dependent way.

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/synthetic_data}
  \caption{\textbf{Recovering known dynamic first-order correlations
      from synthetic data.}  Each panel displays the average
    correlations between the vectorized upper triangles of the
    recovered correlation matrix at each timepoint and either the true
    underlying correlation at each timepoint or a reference
    correlation matrix.  (The averages are taken across 10 different
    randomly generated synthetic datasets of the given category.)
    Error ribbons denote 95\% confidence intervals (taken across
    datasets). Different colors denote different kernel shapes, and
    the shading within each color family denotes the kernel width
    parameter.  For a complete description of each synthetic dataset,
    see \textit{Synthetic data: simulating dynamic first-order
      correlations}.  \textbf{a. Constant correlations.}  These
    datasets have a stable (unchanging) underlying correlation matrix.
    \textbf{b. Random correlations.}  These datasets are generated
    using a new independently drawn correlation matrix at each new
    timepoint.  \textbf{c. Ramping correlations.}  These datasets are
    generated by smoothly varying the underlying correlations between
    the randomly drawn correlation matrices at the first and last
    timepoints.  The left panel displays the correlations between the
    recovered dynamic correlations and the underlying ground truth
    correlations.  The middle panel compares the recovered
    correlations with the \textit{first} timepoint's correlation
    matrix.  The right panel compares the recovered correlations with
    the \textit{last} timepoint's correlation matrix.
    \textbf{d. Event-based correlations.}  These datasets are each
    generated using five randomly drawn correlation matrices that each
    remain stable for a fifth of the total timecourse.  The left panel
    displays the correlations between the recovered dynamic
    correlations and the underlying ground truth correlations.  The
    right panels compare the recovered correlations with the
    correlation matrices unique to each event.}
  \label{fig:synthetic}
\end{figure}

\subsubsection*{Recovering dynamic higher-order correlations}
Following our approach to evaluating our ability to recover known
dynamic first-order correlations from synthetic data, we generated an
analogous second set of synthetic datasets that we designed to exhibit
known dynamic first-order \textit{and} second-order correlations (see
\textit{Synthetic data: simulating dynamic higher-order
  correlations}).  We generated a total of 40 datasets that varied in
how the first-order and second-order correlations changed over time.
We then repeatedly applied Equation~\ref{eqn:timecorr} using the
overall best-performing kernel from our first-order tests (a Laplace
kernel with a width of 20; Fig.~\ref{fig:synthetic}) to assess how
closely the recovered dynamic correlations matched the dynamic
correlations we had embedded into the datasets.

Overall, we found that we could reliably recover both first-order and
second-order correlations from the synthetic data
(Fig.~\ref{fig:higher_order_sims}).  When the correlations were stable
for longer intervals, or changed gradually (constant, ramping, and
event datasets), recovery performance was relatively high, and we were
better able to recover dynamic first-order correlations than
second-order correlations.  We expected that this would happen, given
that errors in our estimation procedure at lower orders necessarily
propagate to higher orders (since lower-order correlations are used to
estimate higher-order correlations).  Interestingly, we also found
that when the correlations were particularly \textit{un}stable (random
datasets), we better recovered second-order correlations.

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/higher_order_sims}
  \caption{\textbf{Recovery of simulated first-order and second-order dynamic correlations.}  Each panel displays the average correlations
    between the vectorized upper triangles of the recovered
    correlation matrix and the simulated correlation matrix at each
    timepoint and for each synthetic dataset.  (The averages are taken across 10 different
    randomly generated synthetic datasets of the given category.)
    Error ribbons denote 95\% confidence intervals (taken across
    datasets).  For a complete description of each synthetic dataset,
    see \textit{Synthetic data: simulating dynamic higher-order
      correlations}.  All results in this figure used a Laplace kernel
    (width~$=20$).  \textbf{a. Constant correlations.}
    These first and second order dynamic correlation were
    simulated from datasets have a stable (unchanging) underlying correlation
    matrix.  \textbf{b. Random correlations.}  These first and second order dynamic correlation were
    simulated from datasets are
    generated using a new independently drawn correlation matrix at
    each new timepoint.  \textbf{c. Ramping correlations.}  These first and second order dynamic correlation were
    simulated from
    datasets are generated by smoothly varying the underlying
    correlations between the randomly drawn correlation matrices at
    the first and last timepoints and displays the
    correlations between the recovered dynamic correlations and the
  simualted correlation first and second order correlation matrices.   \textbf{d. Event-based correlations.}  These first and second order dynamic correlation were
    simulated from datasets are
    each generated using five randomly drawn correlation matrices that
    each remain stable for a fifth of the total timecourse.  }
  \label{fig:higher_order_sims}
\end{figure}

Taken together, our explorations using synthetic data indicated that
we are able to partially, but not perfectly, recover ground truth
dynamic first-order and second-order correlations.  This suggests that
our modeling approach provides a meaningful (if noisy) estimate of
high-order correlations.  We next turned to analyses of human fMRI data
to examine whether the recovered dynamics might reflect the dynamics
of human cognition during a naturalistic story-listening task.

In brief, we computed timeseries of dynamic high-order correlations
that were similar across participants in each of two randomly assigned
groups: a training group and a test group.  We then trained
classifiers on the training group's data to match each sample from the
test group with a stimulus timepoint.  Each classifier comprised a
weighted blend of neural patterns that reflected up to
$n^\mathrm{th}$-order dynamic correlations (see \textit{Feature
  weighting and testing}, Fig.~\ref{fig:pipeline}).  We repeated this process for
$n \in \left\{ 0, 1, 2, ..., 10 \right\}$.  Our examinations of
synthetic data suggested that none of the kernels we examined were
``universal'' in the sense of optimally recovering underlying
correlations regardless of the temporal structure of those
correlations.  We found a similar pattern in the (real) fMRI data,
whereby different kernels yielded different decoding accuracies, but
no single kernel emerged as the clear ``best.''  In our analyses of
neural data, we therefore averaged our decoding results over a variety
of kernel shapes and widths in order to identify results that were
robust to specific kernel parameters (see \textit{Identifying robust
  decoding results}).

Our approach to estimating dynamic high-order correlations entails
mapping the high-dimensional feature space of correlations
(represented by a $T$ by
$\mathcal{O}(K^2)$ matrix) onto a lower-dimensional feature space
(represented by a $T$ by $K$ matrix).
We carried out two sets of analyses that differed in how this mapping
was computed.  The first set of analyses used PCA to find a
low-dimensional embedding of the original dynamic correlation matrices
(Fig.~\ref{fig:decoding}a,b).  The second set of analyses
characterized correlations in dynamics of each feature's eigenvector
centrality, but did not preserve the underlying activity dynamics
(Fig.~\ref{fig:decoding}c,d).

An open question not addressed by our study pertains to how different
structures integrate incoming information with different time
constants.  For example, one line of work suggests that the cortical
surface comprises a structured map such that nearby brain structures
process incoming information at similar timescales.  Low-level sensory
areas integrate information relatively quickly, whereas higher-level
regions integrate information relatively slowly~\citep{BaldEtal17,
  HassEtal08, HassEtal15, HoneEtal12a, LernEtal11, LernEtal14,
  ChieHone19}.  A similar hierarchy appears to play a role in
predicting future events~\citep{LeeEtal20}.  Other
related work in human and mouse brains indicates that the temporal
response profile of a given brain structure may relate to how strongly
connected that structure is with other brain
areas~\citep{FallEtal19}.  Further study is needed to understand the
role of temporal integration at different scales of network
interaction, and across different anatomical structures. Importantly,
our analyses do not speak to the physiological basis of higher-order
dynamics, and could reflect nonlinearities, chaotic patterns,
non-stationarities, or multistability, etc. However, our decoding analyses do indicate
that higher-order dynamics are
consistent across individuals, and therefore unlikely to be driven by
non-stimulus-driven dynamics which are unlikely to be similar across
individuals.

\section*{Methods}
Our general approach to efficiently estimating high-order dynamic
correlations comprises four general steps (Fig.~\ref{fig:methods}).
First, we derive a kernel-based approach to computing dynamic pairwise
correlations in a $T$ (timepoints) by $K$ (features) multivariate
timeseries, $\mathbf{X}_0$.  This yields a $T$ by $\mathcal{O}(K^2)$
matrix of dynamic correlations, $\mathbf{Y}_1$, where each row
comprises the upper triangle and diagonal of the correlation matrix at a single
timepoint, reshaped into a row vector (this reshaped vector is
$\left( \frac{K^2 - K}{2} + K \right)$-dimensional).  Second, we apply a dimensionality
reduction step to project the matrix of dynamic correlations back onto
a $K$-dimensional space.  This yields a $T$ by $K$ matrix,
$\mathbf{X}_1$, that reflects an approximation of the dynamic
correlations reflected in the original data.  Third, we use repeated
applications of the kernel-based dynamic correlation step to
$\mathbf{X}_n$ and the dimensionality reduction step to the resulting
$\mathbf{Y}_{n+1}$ to estimate high-order dynamic correlations.  Each
application of these steps to a $T$ by $K$ time series $\mathbf{X}_n$
yields a $T$ by $K$ matrix, $\mathbf{X}_{n+1}$, that reflects the
dynamic correlations between the columns of $\mathbf{X}_n$.  In this
way, we refer to $n$ as the \textit{order} of the timeseries, where
$\mathbf{X}_0$ (order 0) denotes the original data and $\mathbf{X}_n$
denotes (approximated) $n^\mathrm{th}$-order dynamic correlations
between the columns of $\mathbf{X}_0$.  Finally, we use a
cross-validation--based decoding approach to evaluate how well
information contained in a given order (or weighted mixture of orders)
may be used to decode relevant cognitive states.  If including a given
$\mathbf{X}_n$ in the feature set yields higher classification
accuracy on held-out data, we interpret this as evidence that the
given cognitive states are reflected in patterns of
$n^\mathrm{th}$-order correlations.

All of the code used to produce
the figures and results in this manuscript, along with links to the
corresponding datasets, may be found at
\href{https://github.com/ContextLab/timecorr-paper}{github.com/ContextLab/timecorr-paper}.
In addition, we have released a Python toolbox for computing dynamic
high-order correlations in timeseries data; our toolbox may be found
at
\href{https://timecorr.readthedocs.io/}{timecorr.readthedocs.io}.

@@

Given a kernel function $\kappa_t(\cdot)$ for timepoint $t$,
evaluated at timepoints $\tau \in \left[ 1, ..., T \right]$, we
can update the static correlation formula in Equation~\ref{eqn:corr}
to estimate the \textit{instantaneous correlation} at timepoint $t$:
\begin{align}
  \mathrm{timecorr}_{\kappa_t}\left(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)\right) &= \frac{\sum_{\tau=1}^T \left( \mathbf{X}(\tau, i) -
                                       \widetilde{\mathbf{X}}_{\kappa_t}(\cdot,
                                                                                        i) \right)
                                 \left( \mathbf{X}(\tau, j) -
                                        \widetilde{\mathbf{X}}_{\kappa_t}(\cdot,
                                                                                        j)\right)}
              {\sqrt{\sum_{\tau=1}^T
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot,
                                                                                        i))
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, j))}},~\mathrm{where}\label{eqn:timecorr}\\
  \widetilde{\mathbf{X}}_{\kappa_t}(\cdot, k) &= \sum_{\tau=1}^T
                       \kappa_t(\tau)\mathbf{X}(\tau, k),\\
  \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, k)) &= \sum_{\tau=1}^T
                                                  \left(
                                                  \mathbf{X}(\tau, k) -
                            \widetilde{\mathbf{X}}_{\kappa_t}(\cdot, k) \right)^2.
\end{align}
Here
$\mathrm{timecorr}_{\kappa_t}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot,
j))$ reflects the correlation at time $t$ between columns $i$ and $j$
of $\mathbf{X}$, estimated using the kernel $\kappa_t$.  We evaluate
Equation~\ref{eqn:timecorr} in turn for each pair of columns in
$\mathbf{X}$ and for kernels centered on each timepoint in the
timeseries, respectively, to obtain a $T$ by $K$ by $K$ timeseries of
dynamic correlations, $\mathbf{Y}$.  For convenience, we then reshape
the upper triangles and diagonals of each timepoint's symmetric correlation matrix into a row
vector to obtain an equivalent $T$ by $\left( \frac{K^2 - K}{2} + K \right)$ matrix.

\subsubsection*{Dynamic inter-subject functional connectivity (DISFC)}
Equation~\ref{eqn:timecorr} provides a means of taking a single
observation matrix, $\mathbf{X}_n$ and estimating the dynamic
correlations from moment to moment, $\mathbf{Y}_{n+1}$.  Suppose that
one has access to a set of multiple observation matrices that reflect
the same phenomenon.  For example, one might collect neuroimaging data
from several experimental participants, as each participant performs
the same task (or sequence of tasks).  Let $\mathbf{X}_n^1$,
$\mathbf{X}_n^2$, ..., $\mathbf{X}_n^P$ reflect the $T$ by $K$
observation matrices ($n = 0$) or reduced correlation matrices
($n > 0$) for each of $P$ participants in an experiment.  We can use
\textit{inter-subject functional connectivity}~\citep[ISFC;
][]{SimoEtal16, SimoChan20} to compute the stimulus-driven correlations reflected
in the multi-participant dataset at a given timepoint $t$ using:
\begin{align}
\bar{\mathbf{C}}(t) = M\left(R\left(\frac{1}{2P} \sum_{p=1}^P
  Z\left(\mathbf{Y}_{n+1}^p(t)\right)^\top + Z\left(\mathbf{Y}_{n+1}^p(t)\right)\right)\right),\label{eqn:disfc}
\end{align}
where $M$ extracts and vectorizes the upper triangle and diagonal of a symmetric
matrix, $Z$ is the Fisher $z$-transformation~\citep{Zar10}:
\begin{align}
Z(r) = \frac{\log(1+r) - \log(1-r)}{2},
\end{align}
$R$ is the inverse of $Z$:
\begin{align}
R(z) = \frac{\exp(2z - 1)}{\exp(2z + 1)},
\end{align}
and $\mathbf{Y}_{n+1}^p(t)$ denotes the correlation matrix at timepoint $t$
(Eqn.~\ref{eqn:timecorr}) between each column of $\mathbf{X}_n^p$ and each
column of the average $\mathbf{X}_n$ from all \textit{other}
participants, $\bar{\mathbf{X}}_n^{ \backslash p}$:
\begin{align}
  \bar{\mathbf{X}}_n^{ \backslash p} = \frac{1}{P-1}\sum_{q \in
  \backslash p} \mathbf{X}_n^q,
\end{align}
where $\backslash p$ denotes the set of all participants other than
participant $p$. In this way, the $T$ by $\left( \frac{K^2 - K}{2} + K \right)$ DISFC
matrix $\bar{\mathbf{C}}$ provides a time-varying extension of the ISFC
approach developed by \cite{SimoEtal16}.

@@

Applying dimensionality reduction algorithms to $\mathbf{Y}$ yields an
$\mathbf{X}$ whose columns reflect weighted combinations (or nonlinear
transformations) of the original columns of $\mathbf{Y}$.  This has
two main consequences.  First, with each repeated dimensionality
reduction, the resulting $\mathbf{X}_n$ has lower and lower fidelity
(with respect to what the ``true'' $\mathbf{Y}_n$ might have looked
like without using dimensionality reduction to maintain tractability).
In other words, computing $\mathbf{X}_n$ is a lossy operation.
Second, whereas each column of $\mathbf{Y}_n$ may be mapped
directly onto specific pairs of columns of $\mathbf{X}_{n-1}$, the
columns of $\mathbf{X}_n$ reflect weighted combinations and/or
nonlinear transformations of the columns of $\mathbf{Y}_n$.  Many
dimensionality reduction algorithms are invertible (or approximately
invertible).  However, attempting to map a given $\mathbf{X}_n$ back
onto the original feature space of $\mathbf{X}_0$ will usually require
$\mathcal{O}(TK^{2n})$ space and therefore becomes intractable
as $n$ or $K$ grow large.

\subsubsection*{Graph measure approaches to computing
  $\mathbf{X}_n$}
The above dimensionality reduction approaches to approximating a given
$\mathbf{Y}_n$ with a lower-dimensional $\mathbf{X}_n$ preserve a
(potentially recombined and transformed) mapping back to the original
data in $\mathbf{X}_0$.  We also explore graph measures that instead
characterize each feature's relative \textit{position} in the broader
network of interactions and connections.  To illustrate the
distinction between the two general approaches we explore, suppose a
network comprises nodes $A$ and $B$, along with several other nodes.  If $A$ and $B$ exhibit
uncorrelated activity patterns, then by definition the functional connection
(correlation) between them will be close to 0.
However, if $A$ and $B$ each interact with \textit{other} nodes in similar ways, we
might attempt to capture those similarities between $A$'s and $B$'s
interactions with those other members of the network.

@@

Because $\mathbf{X}_n$ has the same shape as the original data
$\mathbf{X}_0$, approximating $\mathbf{Y}_n$ with a lower-dimensional
$\mathbf{X}_n$ enables us to estimate high-order dynamic correlations
in a scalable way.  Given a $T$ by $K$ input matrix, the output of
Equation~\ref{eqn:timecorr} requires $\mathcal{O}(TK^2)$ space to
store.  Repeated applications of Equation~\ref{eqn:timecorr} (i.e.,
computing dynamic correlations between the columns of the outputted
dynamic correlation matrix) each require exponentially more space; in
general the $n^\mathrm{th}$-order dynamic correlations of a $T$ by $K$
timeseries occupies $\mathcal{O}(TK^{2^n})$ space.  However, when we
approximate or summarize the output of Equation~\ref{eqn:timecorr} with a $T$ by
$K$ matrix (as described above), it becomes feasible to compute even
very high-order correlations in high-dimensional data.  Specifically,
approximating the $n^\mathrm{th}$-order dynamic correlations of a $T$
by $K$ timeseries requires only $\mathcal{O}(TK^2)$ additional space--
the same as would be required to compute first-order dynamic
correlations. In other words, the space required to store $n+1$
multivariate timeseries reflecting up to $n^\mathrm{th}$ order
correlations in the original data scales linearly with $n$ using our
approach (Fig.~\ref{fig:methods}).

\subsection*{Data}
We examined two types of data: synthetic data and human functional
neuroimaging data.  We constructed and leveraged the synthetic data to
evaluate our general approach~\citep[for a related validation approach
see][]{ThomEtal18}.  Specifically, we tested how well
Equation~\ref{eqn:timecorr} could be used to recover known dynamic
correlations using different choices of kernel ($\kappa$;
Fig.~\ref{fig:kernels}), for each of several synthetic datasets that
exhibited different temporal properties. We also simulated
higher-order correlations and tested how well
Equation~\ref{eqn:timecorr} could recover these correlations using the
best kernel from the previous synthetic data analyses.  We then applied our
approach to a functional neuroimaging dataset to test the hypothesis
that ongoing cognitive processing is reflected in high-order dynamic
correlations.  We used an across-participant classification test to
estimate whether dynamic correlations of different orders contain
information about which timepoint in a story participants were
listening to.

\subsubsection*{Synthetic data: simulating dynamic first-order correlations}
We constructed a total of 40 different multivariate timeseries,
collectively reflecting a total of 4 qualitatively different patterns
of dynamic first-order correlations (i.e., 10 datasets reflecting each type of
dynamic pattern).  Each timeseries comprised 50 features (dimensions)
that varied over 300 timepoints.  The observations at each timepoint
were drawn from a zero-mean multivariate Gaussian distribution with a
covariance matrix defined for each timepoint as described below.  We
drew the observations at each timepoint independently from the draws
at all other timepoints; in other words, for each observation
$s_t \sim \mathcal{N}\left(\mathbf{0}, \mathbf{\Sigma}_t\right)$ at
timepoint $t$, $p(s_t) = p(s_t | s_{\backslash t})$.
